Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis

### 2015-05-11 paper/main.tex
take 1, Agusti paper

\documentclass[smallextended]{svjour3}
\input{TexOpts.tex}
\begin{document}
\title{Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis}
\author{Edward A. Roualdes \and Simon Bonner \and Thomas D. Whitney \and James D. Harwood}
\institute{E.A. Roualdes \and S. Bonner
        \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu} \and
        T.D. Whitney 
        \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA \emph{Present address: Warnell School of Forestry and Natural Resources, University of Georgia, Athens, GA 30602, USA}
        \and
        J.D. Harwood
        \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA
}

@@

The indices most commonly used to describe a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them.  A quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.  These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.  Other authors \citep{Davey:2013,King:2010} are using the Monte Carlo based method of \citet{Agusti:2003}, but this also neglects to take into account multiple prey species across multiple time points.  We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.

@@

\citet{Manly:2002} recognized the need for more formal statistical inference and proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for formal hypothesis testing.  Much of this work however focues on general resource selection and doesn't directly address the problems of food selection.  Our model, while focusing on predators' food selection, not general resource selection, offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:2002}, but also provides meaningful single number summaries of the predator's dietary preferences.  Thus, we are proposing here a model that 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers formal statistical justification similar to GLMs, 3) models multiple prey species across multiple time points, and 4) estimates parameters of interest even when exact counts of prey species eaten are not fully observed.

@@

Our model is based on two Poisson distributions with parameters describing the rates at which the predator consumes prey species and the rate at which the predator encounters prey species.  Because the Poisson distirbution is a member of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.  When exact counts are unobserved, we instead rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.

@@

Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.  For instance, suppose interest lies in the predator's dietary preferences with respect to prey species $s_1,s_2$ across $3$ time periods.  If it is believed that hypothesis $4.$ best describes the predator's eating, i.e.\ that the prey species are equally preferred, such that $s_1,s_2$ can be thought of as one prey, but that this preference changes across time, then

\begin{align*}
  H_0&: \lambda_t = c_t \gamma_t, \forall t \\  
  H_1&: \lambda_{st} = c_{st}\gamma_{st}
\end{align*}

would test, statistically, if the data provide evidence for such a belief beyond the model that treats each prey species $s_1$ and $s_2$ individually across the $3$ time points.  Similarly,

@@

would test if $P$'s eating varies by prey species, but not by time, against the most parameter rich model $c_{st}$.  The above two tests should be performed before any of the simpler methods are tested.

@@

All the methods mentioned here, whether they be others' methods or our model, are susceptible to a number of issues that must be handled as best as possible by the investigator.  The most obvious are sampling issues of prey and predator species of interest.  It is almost impossible to know how well one's data represents the habitat and the availability of the prey of interest.  Variables not recorded when collecting data could have influenced the relative abundance of species that were caught.  And yet this is just the beginning.  Other practical challenges of analysing the data once in the lab provide immense opportunity for data to be comprimised.  We therefore, in this paper, focused solely on the methodologies themselves, assuming the best of possible scenarios brought your data to our model.

The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002},  produced one number summaries that rarely considered more than one prey species of interest, did not take into account changes across time, and had minimal statistical justification.  Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.  We feel these indices missed the point; an index is only as good as it is understood to estimate and model our world of observable data.

A more recent and notable method to esimate predator's preferences uses a Monte Carlo, or bootstrap to be more accurate, approach \citep{Agusti:2003}.  This method deserves credit for its lack of distributional assumptions, and yet it still misses the benefits of formally modelling multiple prey species and multiple time points.  Further, it should be noted that this Monte Carlo approach though creative and widely applicable, easily confuses a key statistical idea: re-sampling does not in any way increase one's sample size or the information contained there within.

The model presented here is a formal statistical model, which enables hypothesis testing of multiple prey species across an array of time points.  When appropriately fit to data our methods, via a likelihood ratio test, offer greater statistical power than any index or distribution-free Monte Carlo methods do.  Robust simulations provided insight into the statistical properties of our model.  Moreover, we can estimate parameters of interest even when the predator's gut contents are not fully observed.

Throughout this paper we assume the observed data were generated from Poisson distributions.  This enabled us to use the expectation-maximisation algorithm to fit our model when only incomplete data is observed.  Of course, other distributional assumptions could be made in the case of fully observed count data, but it is not clear if this is true in the case of the unobserved count data.  For instance, consider simply replacing the Poisson distribution with a distribution that has more than one parameter.  Since, in the case of non-count data only $0$s and $1$s are observed, a distribution with more than one parameter would be over-identified.  The only other option is to specify a distribution that better fits the observed data than the Poisson distribution, but here one needs to be careful about the underyling processes that are being modelled.

{\color{red} So what does our model offer ecologists? Just a quick paragraph.}

Further developments of our model could take into account other environmental variables that might affect a predator's eating habits, such as rain or temperature.

\begin{acknowledgements}
The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.  Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.  
\end{acknowledgements}
% \section{Data Accessibility}
% \label{sec:accessibility}

@@

>>>>

\documentclass[smallextended]{svjour3}
\input{TexOpts.tex}
\begin{document}
\title{Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis}
\author{Edward A. Roualdes \and Simon J. Bonner \and Thomas D. Whitney \and James D. Harwood}
\institute{E.A. Roualdes \and S. J. Bonner
        \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu} \and
        T.D. Whitney 
        \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA \emph{Present address: Warnell School of Forestry and Natural Resources, University of Georgia, Athens, GA 30602, USA}
        \and
        J.D. Harwood
        \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA
}

@@

The indices most commonly used to describe a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them.  A quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.  These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.  Other authors are using the Monte Carlo based method of \citet{Agusti:2003} \citep[see][]{Davey:2013,King:2010}, but this is also unable to take into account multiple prey species across multiple time points.  We propose an intuitive statistical model to estimate and statistically test differences in a predator's prey preferences across an array of time points and between multiple prey species.

@@

\citet{Manly:2002} recognized the need for more formal statistical inference and proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for formal hypothesis testing.  Much of this work however focuses on general resource selection and doesn't directly address the problems of food selection.  Our model focuses specifically on the problem of predators' food selection, offers formal inference and hypothesis testing similar to that of GLMs \citet{Manly:2002}, and provides meaningful single number summaries of the predators' food preferences.  The model we are proposing 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers formal statistical justification similar to GLMs, 3) accommodates data collected from multiple prey species across multiple time points, and 4) provides parameters estimates even when exact counts of prey species eaten are not fully observed.

@@

Our model assumes that numbers of each prey species captured and the numbers consumed by an individual predator follow independent Poisson distributions with separate rates that might possibly vary between species and over time.  We are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.  When exact counts are unobserved, we instead rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.

@@

Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.  For instance, suppose interest lies in the predator's dietary preferences with respect to prey species $s_1$ and $s_2$ across three time periods.  The hypothesis that the predator has the same preference for the two prey species can be tested with

\begin{align*}
  H_0&: \lambda_t = c_t \gamma_t, \forall t \\  
  H_1&: \lambda_{st} = c_{st}\gamma_{st}.
\end{align*}

The alternative here allows for the predators preferences to vary independently for the two species and across the three time points. Similarly, the hypotheses

@@

would test if the predator's eating varies by prey species, but not by time, against the most parameter rich model $c_{st}$.  The above two tests should be performed before any of the simpler methods are tested.

@@

The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002}, produced one number summaries that rarely considered more than one prey species, did not take into account changes across time, and had minimal statistical justification.  Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.  Further, these indices could not handle unobserved count data.

A more recent method, by \citet{Agusti:2003}, esimates predator's preferences with unobserved count data using a Monte Carlo approach, the bootstrap to be more accurate.  This method deserves credit for its attempt to do away with distributional assumptions, however we find that the lack of a distribution causes some unresolved problems.  \citet{Agusti:2003} suggest resampling the data to estimate proprotions of prey species present in the mesocosms\footnote{Mesocosm was defined by \citet{Agusti:2003} and is there assumed representative of the predator's habitat.} and in predators' guts.  A seemingly intuitive null hypothesis to test random feeding states that pairs of proportions, relative to a given prey species, are equal.  For instance, if the proportion of prey species $j$ in the mesocosm is denoted $\pi_j$, and $p_j$ denotes the proportion of prey species $j$ in the predator's gut, then their null hypothesis is written $\pi_j = p_j, \, \forall j$.  Since the $\pi_j$s depend on the unknown number of prey species consumed, this null hypothesis is not easily tested.  Even if one assumed known the distribution of prey species consumed, this assumed distribution would depend on unkown parameters, thus only complicating any derived null hypothesis further.  Exactly what \citet{Agusti:2003} are testing in their paper is unclear.

The model presented here is a formal statistical model, which enables clear and intuitive hypothesis testing of multiple prey species across an array of time points.  Our methods, via a likelihood ratio test, offer statistically powerful conclusions about the hypotheses of interest.  Robust simulations provided insight into the statistical properties of our model.  Moreover, we can estimate parameters of interest even when the predator's gut contents are not fully observed.

Throughout this paper we assume the observed data were generated from underlying Poisson distributions.  Though this assumption is not easily tested, and other factors affecting our model may distort our methods, e.g.\ sampling issues, environmental variables, differential decay rates, we do not believe a better assumption is readily available to handle all the cases we considered.  The Poisson distribution provides reasonable biological interpretation to the available data, data obtained by ecologists interested in prey electivity.  One could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.  This may perform better than our model when full count data are observed.  But when only unobserved count data is available, any distribution with more than one parameter is unidentifiable.  Further, because the Poisson distribution {\color{red} SAY SOMETHIGN}, the EM algorithm is theoretically guaranteed to converge to the parameters of interest {\color{red} CITATION? Hoeting and Givens}.  Future development of our methods might consider more complex distributions to handle variables not presently considered in the case of fully observed count data.

{\color{red} So what does our model offer ecologists? Multiple prey species and time points, contrasts, unobserved count data.  Just a quick paragraph. It should also mention our package.}

@@

\begin{acknowledgements}
The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.  Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.  
\end{acknowledgements}
